Pythran: Bringing OpenMP to a Python Subset

### 2013-02-10 doc/papers/europar2013/paper.tex
Take into account FI's review.

To illustrate this behavior, we used the CPU-bound Buffon's needle algorithm to
estimate the value of $\pi$. It's execution time using a multi-threaded
implementation on 4 cores is $\times 1.46$ the execution time of the sequential
version. GIL contention actually makes the code run slower.

@@

An opposite approach consists in using the host language ---herein Python---
to describe both parts of the system, and to let an automated tool perform the
translation to native code of a part of the application, generally the
computation-intensive one where parallelism has been expressed in some ways.
Thus, developers not familiar with lower level languages or not eager to invest
the additional development time can still benefit from a fair performance boost.
This approach is the subject of many studies that can be classified based on
their compatibility with the host language.

@@

A constraining (from the performance point of view) aspect of the Python
language is its type system. It has several consequences, among which the fact
that each method call is resolved dynamically, including a simple add
operation! It comes at no surprise that many approaches extend the Python
language to add a static typing overlay. Likewise, there are only two types of
integers (64bits integers and multi-precision integers) and one type of floating
point type (double-precision floats) in Python, so using a type with the
appropriate size may lead to significant performance boost.

Cython~\cite{cython2010} is such a Python dialect. It extends the syntax with
typing informations, calls to native functions from third party libraries, and a
limited set of parallelism constructs, among which the possibility to define
parallel loops, but no task parallelism. PLW~\cite{dongara2007} proposes a
similar approach that mixes Python with C, using raw strings to hold the C code
and the parallel directives, that are limited to parallel for loops.
%The numba~\footnote{\emph{cf}.\ \texttt{https://github.com/numba/numba}}
%compiler uses additional type information to generate sequential LLVM bytecode.
PyCUDA and PyOpenCL~\cite{klockner2012} also target accelerators by mixing
Python with kernels embedded as raw strings to target accelerators.

The main issue of these approaches is that they imply to modify, in a more or
less complex way, the original code. They require to invest in a new dialect,
and the long-term preservation of this investment is not ensured. Moreover,
there is no falling back if the code were to be deployed in an environment
where the parallelizing tool/module is not available. For instance, once a code
has been ported to PyCUDA and transformed into CUDA code embedded in Python
strings, there is no way back and the developer has to maintain two versions of
the algorithm.

@@

Most backward-compatible approaches also require to modify the input program,
but they do not extend the Python language, rather restrict it. As a
consequence, they remain compatible with the original language and do not suffer
from the drawbacks of the previous approaches. They also keep the benefit of
existing tools associated to the language.

Copperhead~\cite{copperhead2011} is a functional, data parallel language embedded in
Python. It uses n-uplets, numpy arrays and lists as its core data structure and
prohibits usage of many control-flow operators such as loops, enforcing the use
of the \texttt{map}, \texttt{filter} or \texttt{reduce} intrinsics to exhibit
parallelism. In returns, it can be efficiently compiled to either CUDA or C++
with calls to the Thrust\footnote{\emph{cf}.\
\texttt{http://thrust.github.com/}} library. Python decorators are used to
identify hot-spots that are JIT-compiled to native code.

@@

Pythran is a subset of Python focused on scientific computing. It is implicitly
statically typed and supports most Python constructs except those that involve
introspection (e.g.\ \texttt{getattr}) or runtime compilation (e.g.\
\texttt{eval}). A few standard modules are supported in addition to the core
language (e.g.\ \texttt{math}, \texttt{random}). The \texttt{numpy} module is
currently only partially supported. The associated compiler turns Pythran code,
possibly annotated with OpenMP and a few function type annotations, into C++
code. Its processing is summarized in Figure~\ref{fig:pythran-compiler} and
relies on type inference to remove most dynamic behavior, combined with a
high-level runtime library, \texttt{pythonic++}, to allow a one-to-one mapping
between Python and C++ constructs. The purpose of this section is not to dig
into the internals of the compiler, but rather to focus on the impact of the
parallelism layer, especially on the runtime library.

@@

Using shared references simplifies memory management, but the counterpart is an
extra atomic operation for each copy. As it limits parallelism, reducing the
number of copies becomes an important goal. The move semantic introduced in
C++11 avoids a few copies when working on temporary objects, but argument
passing still implies copies. To avoid a reference increment, one can pass
parameters by reference. Using an out-of-scope interprocedural memory effect
analysis, Pythran determines for each argument whether it must be passed using
reference or const reference to prevent this extra cost.

@@

When passing containers from Python to C back and forth, a copy of the whole
data is made to turn the type agnostic, non-contiguous Python data into dense
typed ones. This extra copy implies an extra cost that is not negligible:
Passing two lists of float from Python to C++ requires as much as half the time
to compute the dot product of the same lists directly in Python. Following
Amdahl's law, this transfer cost greatly hinders parallelization, and is a well
known issue.

@@

This work has received fundings from the \textsc{silkan} company and the French
ANR through the CARP project. The authors would like to thank Adrien
\textsc{Merlini} and Alan \textsc{Raynaud} for their work on the C++ runtime,
Albert \textsc{Cohen}, B{\'e}atrice \textsc{Creusillet}, Fabien
\textsc{Dagnat} and Mehdi \textsc{Amini} for their valuable advices.

>>>>

To illustrate this behavior, we used the CPU-bound Buffon's needle algorithm to
estimate the value of $\pi$. It's execution time using a multi-threaded
implementation on 4 cores is $\times 1.46$ the execution time of the sequential
version. GIL contention actually increase the execution time.

@@

An opposite approach consists in using the host language ---herein Python--- to
describe both parts of the system, and to let an automated tool perform the
translation to native code of a part of the application, generally the
computation-intensive one where parallelism has been expressed in some ways.
Thus, developers not familiar with lower level languages or not eager to invest
the additional development time can still benefit from a fair performance boost.
This approach is the subject of many studies that can be classified according to
their compatibility with the host language.

@@

A constraining (from the performance point of view) aspect of the Python
language is its type system. It implies that each method call is resolved
dynamically, even a simple add operation! It comes at no surprise that many
approaches restrain the Python language to add a static typing overlay.  Also,
only two types of integers (64bits integers and multi-precision integers) and
one type of floating point type (double-precision floats) are available in
Python, so using a type with the appropriate size may lead to significant
performance boost.

Cython~\cite{cython2010} is such a Python dialect. It extends the syntax with
typing informations, calls to native functions from third party libraries, and a
limited set of parallelism constructs, such as the possibility to define
parallel loops, but no task parallelism. PLW~\cite{dongara2007} proposes a
similar approach that mixes Python with C, using raw strings to hold the C code
and the parallel directives, that are limited to parallel for loops.
%The numba~\footnote{\emph{cf}.\ \texttt{https://github.com/numba/numba}}
%compiler uses additional type information to generate sequential LLVM bytecode.
PyCUDA and PyOpenCL~\cite{klockner2012} also target accelerators by mixing
Python with kernels embedded as raw strings containing accelerator code.

The main issue of these approaches is that they imply to modify, in a more or
less complex way, the original code. They require to learn a new dialect,
and the long-term preservation of this investment is not ensured. Moreover,
there is no falling back if the code were to be deployed in an environment
where the parallelizing tool/module is not available. For instance, once a code
has been ported to PyCUDA and transformed into CUDA code embedded in Python
strings, there is no way back and the developer has to maintain two versions of
the algorithm.

@@

Most backward-compatible approaches also require to modify the input program.
They do not extend the Python language, but restrict it. As a consequence, they
remain compatible with the original language and do not suffer from the
drawbacks of the previous approaches. They also benefit from existing tools
associated to the language.

Copperhead~\cite{copperhead2011} is a functional, data parallel language
embedded in Python. It uses n-uplets, numpy arrays and lists as its core data
structure and prohibits usage of many control-flow operators such as loops,
enforcing the use of the \texttt{map}, \texttt{filter} or \texttt{reduce}
intrinsics to exhibit parallelism. But it can be efficiently compiled to either
CUDA or C++ with calls to the Thrust\footnote{\emph{cf}.\
\texttt{http://thrust.github.com/}} library. Python decorators are used to
identify hot-spots that are JIT-compiled to native code.

@@

Pythran is a subset of Python designed for scientific computing. It is
implicitly statically typed and supports most Python constructs except those
that involve introspection (e.g.\ \texttt{getattr}) or runtime compilation
(e.g.\ \texttt{eval}). A few standard modules are supported in addition to the
core language (e.g.\ \texttt{math}, \texttt{random}). The \texttt{numpy} module
is currently only partially supported. The associated compiler turns Pythran
code, possibly annotated with OpenMP and a few function type annotations, into
C++ code. Its processing is summarized in Figure~\ref{fig:pythran-compiler} and
relies on type inference to remove most dynamic behavior, combined with a
high-level runtime library, \texttt{pythonic++}, to allow a one-to-one mapping
between Python and C++ constructs. The purpose of this section is not to dig
into the internals of the compiler, but rather to focus on the impact of the
parallelism layer, especially on the runtime library.

@@

Using shared references simplifies memory management, but the counterpart is an
extra atomic operation for each copy. As it limits parallelism, reducing the
number of copies becomes an important goal. The move semantic introduced in
C++11 avoids a few copies when working on temporary objects, but argument
passing still implies copies. To avoid a reference increment, one can pass
parameters by reference. Using an out-of-scope interprocedural memory effect
analysis, Pythran determines for each argument whether it must be passed using
reference or const reference to prevent this overhead.

@@

When passing containers from Python to C back and forth, a copy of the whole
container is made to turn the type agnostic, non-contiguous Python data into
dense typed ones. This extra copy implies an extra cost that is not negligible:
Passing two lists of float from Python to C++ requires as much as half the time
to compute the dot product of the same lists directly in Python. Following
Amdahl's law, this copy overhead greatly hinders the benefits of
parallelization, and is a well known issue.

@@

This work has received fundings from \textsc{silkan} and the French ANR through the
CARP project. The authors thank Adrien \textsc{Merlini} and Alan
\textsc{Raynaud} for their work on the runtime, Albert \textsc{Cohen},
B{\'e}atrice \textsc{Creusillet}, Fabien \textsc{Dagnat}, Mehdi \textsc{Amini},
Fran{\,c}ois \textsc{Irigoin} and Ronan \textsc{Keryell} for their valuable
reviews.
